Investigating soundscapes perception through acoustic scenes simulation

### 2018-07-16 v2/responseToReviewers.tex

\item \emph{In the last sentence of the abstract, the distinction of 'physical descriptors' vs. 'global descriptors'  does not have enough context here to be informative.}

$\rightarrow$ This sentence have be rewritten to be more explicit: "acoustic pressure levels computed for specific sound sources better characterize the appraisal than the acoustic pressure level computed over the overall soundscape."

\item \emph{In section 2.1 (pg. 5), the citations (particularly 27-31) do not seem relevant for justifying the event/texture distinction in the simulator.}

$\rightarrow$ The references to the ASA theory are now removed. The work of McDermott and Nelken and de Cheveigné are still cited as, in our humble opinion, they have some relevance to the matter discussed in this section.

### 2018-07-16 lafaySimulatedScenePerception.tex
Update section 4.3

As asserted by~\cite{aletta2016soundscape}, one of the main goals of soundscape studies is to identify which components of the soundscape influence human perception. Establishing a link between an environment (urban sound scene) and the induced human sensation (calmness) is equivalent to instantiating the mental representation of a specific sound scene (a calm urban sound scene). Several means have been considered in the literature to perform this task.

@@

We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene, see~\cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene, as found by~\cite{gozalo2015relationship,ricciardi2015sound}.

@@

\begin{enumerate}
\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non-ideal soundscapes that are considered as material for the following experiments; The resulting waveforms are available for download \url{https://archive.org/details/soundSimulatedUrbanScene}.
\item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
\item \emph{experiment 2, pleasantness evaluation after modification of the scenes}: the subjects judge the pleasantness of the simulated scenes on a semantic scale as in 1.b, but some scenes are modified beforehand, \ie~some specific sounds classes that are identified as having a significant impact on perceived pleasantness are removed.
\end{enumerate}

@@

The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts written by~\cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} that study how humans discriminate different kind of urban soundscapes.

@@

\begin{itemize}
\item {sound events}: isolated sounds, precisely located in time, whose acoustical characteristics may change with respect to time;
\item {sound textures}: isolated sounds of long duration whose acoustical characteristics are stable with respect to time, as expressed by~\cite{saint1995classification}.
\end{itemize}

@@

Among the 44 subjects, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non-ideal scenes). The 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.  The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available \url{http://soundthings.org/research/urbanSoundScape/XP2014}.

@@

All the subjects perform the experiment on computers with equivalent hardware and software configurations. The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. The stimuli are the scenes obtained in experiment 1.a. The output sound level is the same for all the subjects.

@@

All the subjects perform the experiment simultaneously. They were not allowed to talk to each other during the experiment.

@@

An experimenter is available during the whole duration of the experiment in order to assist subjects and answer questions any questions the subjects may have.

@@

\begin{table}[t]
\centering
\begin{tabular}{c c}
Terms                              & Acronyms                   \\
\hline
Sound level                        & $L$                        \\
Sound level (events)               & $L(E)$                     \\
Sound level (textures)             & $L(T)$                     \\
Average pleasantness (per scene)   & $\mathcal{A}_{scene}$      \\
Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
%                                    &                            \\
% ideal/pleasant                     & i                          \\
% non-ideal/unpleasant               & ni                         \\
% scene ideal/pleasant               & i-scene                    \\
% scene non-ideal/unpleasant         & ni-scene                   \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Acronyms of features used in the analysis of the sensory experiments.}
\label{tab:acronyme}
\end{table}

@@

\item \emph{Influence of sound levels on perceived pleasantness}: the goal is to evaluate if the sound levels impact the perceived pleasantness. To do so, we consider linear correlations between these features and $\mathcal{A}_{scene}$. The Pearson correlation coefficient is used for that purpose.

@@

\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (ideal or non-ideal). To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}. With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of environment $k$ ($k \in \lbrace ideal, non-ideal\rbrace$), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$. For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:

@@

First, in order to ensure the coherence of the data, we check that none of the non-ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of a ideal scene. 4 non-ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes. As a consequence, 36 ideal scenes and 36 non-ideal scenes remain for analysis. Second, we verify that subjects really perceived a difference in terms of pleasantness between ideal and non-ideal scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It indeed appears that the ideal scenes were perceived significantly more pleasant ($p < 0.01$) than the non-ideal scenes for all the participants.

@@

This difference may be explained by the nature of the experimental protocol used in the two studies. As in our study, Guastavino asks the subjects to describe an environment. She asks them to perform this task starting only from their memories, whereas in our case subjects perform the same task using actual sound samples that they can listen to. The fact that subjects in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact. Other studies that considered sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment, as studied by~\cite{lavandier2006contribution}.

@@

Considering $E$ and $T$ separately, it appears that: 1) the rank 5-precision obtained with $E$ is similar to the one obtained with $ET$ ; 2) the rank 5-precision obtained with $T$ is always lower than the one obtained with $E$, by $10\%$ to $15\%$. Those results indicate that the semantic information that is discriminative is mostly carried by the events. Those results are in line with results of~\cite{maffiolo_caracterisation_1999}. As discussed in Section~\ref{sec:simscene_sampleDataSet}, it seems that humans analyze the event scenes which are composed of several sound events in a causal manner, \ie~by identifying the sources.

@@

Concerning the sound levels, the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ and $L(T)$ on the other side, see~Figures~\ref{fig:soundlevelMarkera} and~\ref{fig:soundlevelMarkerd}, respectively. No matter whether all the classes or only the markers are considered, it appears that:

@@

To conclude, the level of ni- markers has a negative influence on the pleasantness for the non-ideal scenes. On the other hand, the level of the ideal markers does not influence the perceived pleasantness for the ideal scenes.

@@

Considering the non markers classes, we can observe on the ideal scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}. It is the first time that an objective feature allows us to define the pleasantness of ideal scenes. This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.

@@

A last group of features is now considered, namely $L_m-L_b$, $L(E)_m-L(E)_b$ and $L(T)_m-L(T)_b$.  These features describe the difference between the markers level and those of the other sound classes, see ~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}. They express the saliency of the markers with respect to the sound mixture.

@@

For the ideal scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$). for the non-ideal scenes no correlation is observed. Thus, for the ideal scenes, it is not the absolute markers level that is important, but their relative level with respect to the other sounds composing the scene. A double perceptual mechanism for the ideal environments can thus be observed:

@@

\begin{itemize}
\item the higher the absolute level of sounds not being ideal markers, the weaker the pleasantness,
\item the higher the relative level of ideal markers is, with respect to the remaining sounds, the higher the pleasantness.
\end{itemize}

@@

The aim here is to validate the hypothesis that the removal of ideal markers and non-ideal markers has a significant effect on the perceived pleasantness. To do so, we perform an analysis of variance (ANOVA). We consider the pleasantness rating of each subject on each scene as a dependent variable, and as independent variables, the type of environment (ideal or non-ideal) and the presence/absence of markers. As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA with interaction is used to evaluate whether there exist a significant difference of perceived pleasantness. The two independent variables are considered as \emph{within-subject} factors. The factors being of only two levels each (type: ideal or non-ideal,  marker: with or without), the sphericity hypothesis does not need to be checked. \emph{Post hoc} analyses are done using the Tukey-Kramer procedure. %\jfp{l'analyse est inutile a mon avis, il n'y a que 2 niveaux pour chaque facteur de l'ANOVA.} \gl{2 niveaux par facteur, mais 4 niveaux en tout}

@@

In this section, we study the scores given by the subject while listening to the several types of scenes, namely the ideal scenes with markers, non-ideal scenes with markers, ideal scenes without markers and non-ideal scenes without markers (Figure~\ref{fig:xp2_Ascene}). The repeated measure ANOVA applied to the pleasantness rating shows a significant effect of the type environment (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).

@@

%\subsection{Discussion}

>>>>

One of the main goals of soundscape studies is to identify which components of the soundscape influence human perception, see~\cite{aletta2016soundscape}. Establishing a link between an environment (urban sound scene) and the induced human sensation (calmness) is equivalent to instantiating the mental representation of a specific sound scene (a calm urban sound scene). Several means have been considered in the literature to perform this task.

@@

We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene, see~\cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene, see~\cite{gozalo2015relationship,ricciardi2015sound}.

@@

\begin{enumerate}
\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non-ideal soundscapes that are considered as material for the following experiments; The resulting waveforms are available for download\footnote{Soundscapes, available online \url{https://archive.org/details/soundSimulatedUrbanScene}}.
\item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
\item \emph{experiment 2, pleasantness evaluation after modification of the scenes}: the subjects judge the pleasantness of the simulated scenes on a semantic scale as in 1.b, but some scenes are modified beforehand, \ie~some specific sounds classes that are identified as having a significant impact on perceived pleasantness are removed.
\end{enumerate}

@@

The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts written that study how humans discriminate different kind of urban soundscape, see~\cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} .

@@

\begin{itemize}
\item {sound events}: isolated sounds, precisely located in time, whose acoustical characteristics may change with respect to time;
\item {sound textures}: isolated sounds of long duration whose acoustical characteristics are stable with respect to time, see~\cite{saint1995classification}.
\end{itemize}

@@

Among the 44 subjects, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non-ideal scenes). The 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.  The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available online\footnote{Available online \url{http://soundthings.org/research/urbanSoundScape/XP2014}}.

@@

All the subjects performed the experiment on computers with equivalent hardware and software configurations. The audio files were played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. The stimuli were the scenes obtained in experiment 1.a. The output sound level was the same for all the subjects.

@@

All the subjects performed the experiment simultaneously. They were not allowed to talk to each other during the experiment.

@@

An experimenter was available during the whole duration of the experiment in order to assist subjects and answer any questions the subjects might have.

@@

\begin{table}[t]
\centering
\begin{tabular}{c c}
Terms                              & Acronyms                   \\
\hline
Sound level                        & $L$                        \\
Sound level (events)               & $L(E)$                     \\
Sound level (textures)             & $L(T)$                     \\
Average pleasantness (per scene)   & $\mathcal{A}_{scene}$      \\
Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
                                   &                            \\
                                   &                           \\
                                   &                          \\
                                   &                     \\
                                   &                    \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Acronyms of features used in the analysis of the experiments.}
\label{tab:acronyme}
\end{table}

@@

\item \emph{Influence of sound levels on perceived pleasantness}: the goal is to evaluate if the sound levels impact the perceived pleasantness. To do so, we consider linear correlations between this feature and $\mathcal{A}_{scene}$. The Pearson correlation coefficient is used for that purpose.

@@

\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (ideal or non-ideal). To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}. With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of environment $k$ (ideal or non-ideal), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$. For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:

@@

First, in order to ensure the coherence of the data, we check that none of the non-ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of a ideal scene. 4 non-ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes. As a consequence, 36 ideal scenes and 36 non-ideal scenes remain for analysis. Second, we verify that subjects really perceived a difference in terms of pleasantness between ideal and non-ideal scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It indeed appears that the ideal scenes were perceived significantly more pleasant ($p < 0.01$) than the non-ideal scenes.

@@

This difference may be explained by the nature of the experimental protocol used in the two studies. As in our study, Guastavino asks the subjects to describe an environment. She asks them to perform this task starting only from their memories, whereas in our case subjects perform the same task using actual sound samples that they can listen to. The fact that subjects in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact. Other studies that considered sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment, see~\cite{lavandier2006contribution}.

@@

Considering $E$ and $T$ separately, it appears that: 1) the rank 5-precision obtained with $E$ is similar to the one obtained with $ET$ ; 2) the rank 5-precision obtained with $T$ is always lower than the one obtained with $E$, by $10\%$ to $15\%$. Those results indicate that the semantic information that is discriminative is mostly carried by the events. Those results are in line with results of~\cite{maffiolo_caracterisation_1999}. As discussed in Section~\ref{sec:simscene_sampleDataSet}, it seems that humans analyze the event scenes which are composed of several sound events in a descriptive manner, \ie~by identifying the sources.

@@

Concerning the sound levels, the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ and $L(T)$ on the other side, see~Figures~\ref{fig:soundlevelMarkera} and~\ref{fig:soundlevelMarkerd}. No matter whether all the classes or only the markers are considered, it appears that:

@@

To conclude, the level of non-ideal markers has a negative influence on the pleasantness for the non-ideal scenes. On the other hand, the level of the ideal markers does not influence the perceived pleasantness for the ideal scenes.

@@

Considering the non markers classes, we can observe on the ideal scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see~Figures~\ref{fig:soundlevelNoisea} and~\ref{fig:soundlevelNoised}. It is the first time that an objective feature allows us to define the pleasantness of ideal scenes. This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.

@@

A last group of features is now considered, namely $L_m-L_b$, $L(E)_m-L(E)_b$ and $L(T)_m-L(T)_b$.  These features describe the difference between the markers level and those of the other sound classes, see~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}. They express the saliency of the markers with respect to the sound mixture.

@@

For the ideal scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$). For the non-ideal scenes no correlation is observed. Thus, for the ideal scenes, it is not the absolute markers level that is important, but their relative level with respect to the other sounds composing the scene. A double perceptual mechanism for the ideal environments can thus be observed:

@@

\begin{itemize}
\item the higher the absolute level of sounds not being ideal markers, the weaker the pleasantness,
\item the higher the relative level of ideal markers compared to the remaining sounds, the higher the pleasantness.
\end{itemize}

@@

The aim here is to validate the hypothesis that the removal of ideal and non-ideal markers has a significant effect on the perceived pleasantness. To do so, we perform an analysis of variance (ANOVA). We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (ideal or non-ideal) and the presence/absence of markers. As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA with interaction is used to evaluate whether there exist a significant difference of perceived pleasantness between the scenes with and without markers. The two independent variables are considered as \emph{within-subject} factors. The factors being of only two levels each (type: ideal or non-ideal,  marker: with or without), the sphericity hypothesis does not need to be checked. \emph{Post hoc} analyses are done using the Tukey-Kramer procedure. %\jfp{l'analyse est inutile a mon avis, il n'y a que 2 niveaux pour chaque facteur de l'ANOVA.} \gl{2 niveaux par facteur, mais 4 niveaux en tout}

@@

In this section, we study the scores given by the subject while listening to the several types of scenes, namely the ideal scenes with markers, non-ideal scenes with markers, ideal scenes without markers and non-ideal scenes without markers (Figure~\ref{fig:xp2_Ascene}). The repeated measure ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the type environment (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).

@@
